Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-09-28 psy_peds_2.Rmd
reformating

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "/Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
```

```{r}
# cut-off dates used in the analysis, these should not be changed
history_start_date <- "2019-01-01"
history_end_date <- "2021-05-31"
pandemic_start_date <- "2020-03-15"
```

@@

files <- readInputFiles( path      = "../Input/09212021_2/", 
                         separator = ",",
                         skip      = 0, 
                         verbose   = FALSE )

@@

### Read the file containing race information for those sites recording this variable
if( raceAvailable == TRUE ){
  race_raw <- read.delim(paste0("../", folder_4ce_files, "/LocalPatientRace.csv"), sep = ",", skip = 0)
}
```

@@

clinical_raw_filter <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic","during_pandemic")) %>% 
  dplyr::filter(cohort  %in% c("AllAdm"))
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation > min_hosp_days))
```

```{r}
all_codes <- clinical_raw_filter %>% 
  left_join( obs_raw,  by = c("patient_num", "cohort", "days_since_admission")) %>%
  filter( concept_type == 'DIAG-ICD10')
```

@@

### New hospitalizations
```{r, fig.width=12, fig.height=12}
new_hospitalizations_per_day <- all_codes %>%
  group_by(patient_num, n_hospitalisation) %>% slice(which.min(date)) %>%
  filter( in_hospital == 1, 
         date < '2021-05-01' &
         date >= '2019-03-01' ) %>%
    group_by( date, weeks, month) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

new_hospitalizations_per_week <- new_hospitalizations_per_day %>%
  filter(weeks >= '2019-03-01') %>% 
  group_by( weeks) %>%
  summarise(count = sum(count))

new_hospitalizations_per_month3 <- new_hospitalizations_per_day %>%
  group_by( month) %>%
  summarise(count = sum(count))

@@

### Total hospitalize patients
```{r, fig.height=12, fig.width=12}
total_hospitalizations_per_day <- all_codes %>%
  group_by(patient_num, date ) %>%
  filter(date < '2021-05-01' &
         date >= '2019-03-01' ) %>%
    group_by( date, weeks, month) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

total_hospitalizations_per_week <- total_hospitalizations_per_day %>% 
  filter(weeks > '2019-03-01') %>% 
    group_by( weeks ) %>%
  summarise(count = sum(count))

total_hospitalizations_per_month <- total_hospitalizations_per_day %>%
  group_by( month ) %>%
  summarise(count = sum(count))

@@

>>>>

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
```

```{r}
# cut-off dates used in the analysis, these should not be changed
history_start_date <- "2019-01-01"
history_end_date <- "2021-05-31"
pandemic_start_date <- "2020-03-15"
start_date_plots <- "2019-03-01"
end_date_plots <- "2021-05-01"
```

@@

files <- readInputFiles( path      = folder_4ce_files, 
                         separator = ",",
                         skip      = 0, 
                         verbose   = FALSE )

@@

### Read the file containing race information for those sites recording this variable
if( raceAvailable == TRUE ){
  race_raw <- read.delim(file.path(folder_4ce_files, "/LocalPatientRace.csv"), sep = ",", skip = 0)
}
```

@@

clinical_raw_filter <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))
```

```{r}
stopifnot(unique(clinical_raw_filter$cohort) == "AllAdm")
all_codes <- clinical_raw_filter %>% 
  left_join( obs_raw,  by = c("patient_num", "days_since_admission")) %>%
  filter( concept_type == 'DIAG-ICD10')
stopifnot(unique(all_codes$in_hospital) == 1)
stopifnot(unique(all_codes$len_hospitalisation > min_hosp_days))
```

@@

### New hospitalizations
```{r, fig.width=12, fig.height=12}
new_hospitalizations_per_day <- all_codes %>%
  group_by(patient_num, n_hospitalisation) %>% slice(which.min(date)) %>%
  filter( in_hospital == 1, 
         date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date, weeks, month) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

new_hospitalizations_per_week <- new_hospitalizations_per_day %>%
  filter(weeks >= start_date_plots) %>% 
  group_by( weeks) %>%
  summarise(count = sum(count))

new_hospitalizations_per_month <- new_hospitalizations_per_day %>%
  group_by( month) %>%
  summarise(count = sum(count))

@@

### Total hospitalize patients
```{r, fig.height=12, fig.width=12}
total_hospitalizations_per_day <- all_codes %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count = n_distinct(patient_num))

total_hospitalizations_per_week <- all_codes %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num))

total_hospitalizations_per_month <- all_codes %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( month ) %>%
  summarise(count = n_distinct(patient_num))

@@

#### Psychiatric related conditions vs. any other condition

##### New admissions

##### Total hospitalizations
We have estimated the total number of hospitalizations in the previous steps:
- total_hospitalization_per_day
- total_hospitalization_per_week
- total_hospitalization_per_month

We will use these counts to estimate the percentage and to compare total counts, vs psychiatric related conditions.

```{r}
total_hospitalizations_per_day_psy <- input_data %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( date ) %>%
  summarise(count_psy = n_distinct(patient_num)) %>%
  left_join( total_hospitalizations_per_day ) %>%
  mutate( period = ifelse( date < pandemic_start_date,
                                  "before_pandemic",
                                 "during_pandemic"), 
                percentage = count_psy * 100 / count)

thpd_psyPer <- total_hospitalizations_per_day_psy %>%
  ggplot(aes(x = date, y = percentage, fill = period)) +
  #geom_line()+
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  labs(y = "Percentage",
       x = "Date (by day)",
       title = "Percentage of patients with mental health related ICD codes (daily)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyPer

thpd_psyCount <- total_hospitalizations_per_day_psy %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = date, y = value, fill = counts, color = counts)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by day)",
       title = "Daily patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

```

We will compare the psychiatric conditions with some other disorders (e.g, digestive, respiratory, etc.)
We will use these counts to estimate the percentage and to compare total counts, vs psychiatric related conditions.

```{r}
total_hospitalizations_per_day_disorder_group <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
        icd_description_category %in% c("Psy","Neoplams","Diseases of the respiratory system","Diseases of the digestive system" ))%>%
    group_by( date, icd_description_category ) %>%
  summarise(counts = n_distinct(patient_num))

thpd_disorderGroup <- total_hospitalizations_per_day_disorder_group %>%
  ggplot(aes(x = date, y = counts, fill = icd_description_category, color = icd_description_category)) +
  #geom_point(alpha=0.2) +
  geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "patient number",
       x = "days",
       title = "Total number of hospitalized patients (daily)") +   
  theme(axis.text.x = element_text(angle=45, vjust = 1, hjust=1), legend.position = "bottom", legend.text = element_text(size=7), legend.title = element_text(size=7) ) +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_disorderGroup
```
